---
title: Differential expression analysis
subtitle: OM5220
author: "pz"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    highlight: tango
    df_print: paged
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, result='hold',fig.width=8,tidy=TRUE, cache=FALSE)
knitr::opts_knit$set(progress=TRUE,verbose=TRUE)
```

```{r libraries, cache=FALSE}
library(data.table)
library(magrittr)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size = 16))
library(DESeq2)
library(ggsci)
library(dplyr)
library(pheatmap)
library(openxlsx)
library(ggrepel)
```



```{r decoder}
decoderFile <- "OM5220_naive_decoder.txt"
decoder.data <- fread(decoderFile) %>% as.data.frame()
decoder.data$HIV <- factor(decoder.data$HIV)
decoder.data$CTL <-factor(decoder.data$CTL)
decoder.data$group.ID <- factor(decoder.data$group)
decoder.data$group <- factor(decoder.data$group)
decoder.data = decoder.data[order(decoder.data$CTL, decoder.data$HIV),]
decoder.data$sample.ID = paste0(decoder.data$RUN_ID,"_",decoder.data$SAMPLE_ID)
row.names(decoder.data) = decoder.data$sample.ID
```


```{r counts}
counts <- read.csv(file = "OM5220_gene_counts.txt", header = TRUE, check.names=FALSE, row.names=1)
decoder.data <- decoder.data[decoder.data$sample.ID %in% colnames(counts),]
counts <- counts[,c(decoder.data$sample.ID)]

if (!identical(colnames(counts), decoder.data$sample.ID)) {
  stop("Column names in counts do not match sample IDs in decoder data")
}
```

# Samples

The following samples were part of this analysis:

```{r samples}
as_tibble(decoder.data[,c("sample.ID", "HIV", "CTL")])
```

# Differential expression analysis {.tabset}

Fitting an interaction term, with the main goal to ask the question, how do HIV+ infected cells respond differently to CTL treatment compared to HIV- cells?

In other words,

(HIV+CTL+ - HIV+CTL-) vs. (HIV-CTL+ - HIV-CTL-)


```{r createDDS}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts, 
                                      colData = decoder.data, 
                                      design = ~ HIV * CTL)

keep <- edgeR::filterByExpr(counts(dds), design = model.matrix(~ HIV * CTL, data = colData(dds)))
dds <- dds[keep,]
dds <- estimateSizeFactors(dds)
vsd <- DESeq2::vst(dds, blind = TRUE)
normCounts = assay(vsd)

dds <- DESeq(dds)

baseMeanPerLvl <- sapply(levels(dds$group), function(lvl) rowMeans( counts(dds, normalized=TRUE)[,dds$group == lvl, drop=F]))
```


```{r}
de.all <- list()
alphaTR = 0.01
lfcThres = 0

de.all[["interaction"]] = results(dds, name="HIVpositive.CTLpresent", alpha=0.01) %>% as.data.frame()
```

## Number of DEGs (table)

The following genes were detected as differentially expressed (adjusted p < `r alphaTR` and abs(log2FoldChange) > `r lfcThres`):

```{r printDEGtable}
de.all.sig <- lapply(de.all, subset, padj < alphaTR & abs(log2FoldChange) > lfcThres)

res =  as.data.frame(sapply(de.all.sig, nrow))
colnames(res) <- paste0("padj<",alphaTR, " & ","abs(log2FoldChange)>",lfcThres)

rel.contrs = 1:nrow(res)

as_tibble(cbind(res,"contrast #"=rel.contrs) %>%  tibble::rownames_to_column(var="contrast"))
```


```{r save_result, eval=T, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE,  fig.width=8.3, fig.height=10}
dir.create("./differential_expression_res/", showWarnings = FALSE, recursive = FALSE, mode = "0777")

lapply(rel.contrs, function(x){
  df <- de.all[[x]] 
  name <- names(de.all)[x]
  samples_in_contrast <- colnames(baseMeanPerLvl)
  wb <- createWorkbook()
  posStyle <- createStyle(fontColour = "#000000", bgFill = "#C6EFCE")
  addWorksheet(wb, stringr::str_trunc(name, width=31))
  forExcel <- merge(baseMeanPerLvl[,samples_in_contrast], df, by="row.names")
  forExcel <- forExcel[order(forExcel$padj),]
  writeData(wb, stringr::str_trunc(name, width=31), forExcel, rowNames=T, keepNA=T) 
  rule <- eval(paste0("$",LETTERS[which(colnames(forExcel) == "padj")+1],paste0("2<",alphaTR)))
  conditionalFormatting(wb, stringr::str_trunc(name, width=31), cols=1:(ncol(forExcel)+1), rows=2:nrow(df), rule=rule,  style = posStyle)
  freezePane(wb, stringr::str_trunc(name, width=31), firstActiveRow = 2)
  setColWidths(wb, stringr::str_trunc(name, width=31), cols = 1:ncol(forExcel), widths = "auto")
  saveWorkbook(wb = wb, file =  paste0("./differential_expression_res/deseq2_OM5220_naive",name,"_padj",alphaTR,".res.xlsx"), overwrite=T)
}) %>% invisible()
```


## Volcano plot (abs(logFC > 0.50))

Genes with effect size log2FoldChange > 0.5 or < -0.5 (vertical dashed lines) and padj < 0.01 (horizontal red line) were colored. Blue points were down regulated genes and red points were up regulated genes. Seleect genes are labeled

```{r volcanoPlot_lfc0.50, fig.width=7, fig.height=5, fig.path='OM5220_pdfs/',  dev=c('png','pdf')}
df <- de.all$interaction
lfcThres = 0.5
alphaThres = 0.01

df$color <- ifelse(df$padj < alphaThres & df$log2FoldChange > lfcThres, "#FF0000", 
                   ifelse(df$padj < alphaThres & df$log2FoldChange < -lfcThres, "#0000FF", 
                          "black"))

xlim <- c(-1, 1) * quantile(abs(df$log2FoldChange[is.finite(df$log2FoldChange)]), probs = 0.99) * 1.1

p <- ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = color)) +
  geom_point(size = 1,  alpha=0.6) +
  scale_color_identity() +
  geom_hline(yintercept = -log10(alphaThres), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = c(-lfcThres, lfcThres), linetype = "dashed", color = "gray") +
  theme_classic(base_size = 18) +
  labs(x = expression("Log"[2]*" Fold Change"),
       y = expression("-log"[10]*" (adj. p-value)"),
       title = "Volcano Plot") +
  theme(plot.background = element_rect(fill = "white"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  scale_x_continuous(limits = xlim) 

key_genes <- subset(df, padj < alphaThres & abs(log2FoldChange) > lfcThres & row.names(df) %in% c("EZH2"))
p + geom_label_repel(data = key_genes, aes(label = row.names(key_genes)), 
                           size = 4,
                           box.padding = unit(0.35, "lines"), 
                           point.padding = unit(0.5, "lines"),
                    min.segment.length = 500, color = "black") + 
  theme(aspect.ratio=1) +
  geom_point(data = key_genes, shape=21, fill="yellow", color="yellow", alpha=0.8)

```



# Session Info
```{r session, message=FALSE, warning=FALSE, cache=FALSE,echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```